A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-23 latex/main.tex
Update phylo backbones section

\subsubsection*{Comparison with UShER}
To facilitate comparison with the sc2ts ARG, we first converted the Viridian 
UShER tree from MAT~\cite{McBroome2021} to tskit format, which has similar
properties but is more general 
(see the \nameref{sec:args} section).
We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
and converted the tree topology to tskit node and edge table 
descriptions. We then downloaded the protobuf formatted
tree (tree.all\_viridian.202409.pb.gz), exported the mutations to 
nucleotide format using \texttt{matUtils summary --translate}
(usher version 0.6.3), and translated
these mutations into tskit site and mutation tables. 
The resulting tszip file 
(usher\_viridian\_v1.0.trees.tsz) contains 5,345,019 nodes, 3,364,841
mutations and requires 29MB of storage space.
We validated that the data was correctly encoded by comparing the 
variant data for the UShER tskit tree sequence against our alignments.
There was an exact match between the non-ambiguous nucleotide calls 
at 27,469 sites. 
% Usher tree has 27,508 vs sc2ts 29,893, or 2385 fewer sites 
At the remaining 39 sites for which the UShER tree 
had mutations there were a significant 
number of gap characters (deletion calls), suggesting that the mismatches
could be due to differences in the alignments used as input to UShER.
See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/validate\_usher\_tree.ipynb}{validate\_usher\_tree.ipynb}
for details.

We then computed intersection ARGs for
sc2ts (sc2ts\_viridian\_inter\_v1.2.trees.tsz) and
UShER (usher\_viridian\_inter\_v1.2.trees.tsz)
by simplifying \cite{Kelleher2018efficient,Wong2024}
both to the 2,475,418 samples and 27,507 sites that are shared
(the UShER tree has mutations at 27,508 sites and sc2ts ARG has 29,893).
As there was no documented means of deriving node dates from
the UShER tree, and we wish to visually compare
the phylogenetic backbones,
we also dated the intersection tree by fixing
the sample dates using the ``Date\_tree'' metadata field,
and using tsdate (version 0.2.3) to date internal nodes.
We did not date the main UShER tree in this way because
of the time-travelling samples present.

@@

\subsubsection*{Comparison of phylogenetic backbones}

We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on the 1473 Pango lineages that have perfect ``origination events'' 
as described in \ref{sec:supplement_pango_lineages}, and
omit Pango designations starting with ``X''
to avoid recombinants, leaving 1329 Pango lineages.
To ensure that each sample is a tip (rather than an internal node)
in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 lineages
which are singleton descendants of a recombination event in the ARG
(representing potentially missed Pango recombinant lineages),
as well as 43 samples that are descendants of the recombinant XBB
but which do not start with an ``X''.
This results in a simplified ARG of 1282 samples,
each sample having a different Pango designation.

The simplified ARG contains only two recombination nodes, which
correspond to those on the far right of the genome identified as
artefactual in the FIXME Xref \ref{sec:spurious-ancestrally-important-recombinants} section.
We therefore compare the Usher topology against the first tree in the ARG (covering
positions 1-26858; the majority of the genome). To reduce plot density, we consider
only those Pango lineages with greater than 10 samples, and to aid visualization,
we divide resulting tree into 4 parts: a BA.5 subtree of 187 Pangos,
a Delta subtree of 104 Pangos, a BA.2 subtree (excluding BA.5 descendants) of 118 Pangos,
and a base tree (i.e. excluding all descendants of Delta and BA.2) of 405 Pangos,
Figures \ref{fig:tanglegram_base}-\ref{fig:tanglegram_ba5} show these trees compared
against the equivalent dated UShER topologies.
The order of tips in each plot was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
See \href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/supp\_pdf-cophylogeny-tanglegrams.ipynb}{supp\_pdf-cophylogeny-tanglegrams.ipynb}
for details and the pipeline used.

Inspection of Figures~\ref{fig:tanglegram_base}-\ref{fig:tanglegram_ba5} reveals that the majority of resolvable nodes,
including most high-level Scorpio groupings,
are shared between the ARG and the UShER tree (magenta nodes), and there is extremely close (usually perfect) agreement on topological relations between the Pango lineages.
In the few cases where the trees differ, this is often in terms of degree of resolution of polytomies.
In these cases, sometimes the sc2ts ARG shows greater resolution,
as in the base of the BA.2 subtree, and
sometimes the UShER tree shows greater resolution, as in the close relatives of Beta (B.1.351).

>>>>

\subsubsection*{Conversion of UShER tree to tskit}
To facilitate comparison with the sc2ts ARG, we first converted the Viridian 
UShER tree from MAT~\cite{McBroome2021} to tskit format.
We first downloaded the tree in JSON format (tree.all\_viridian.202409.jsonl)
and converted the tree topology to tskit node and edge table 
descriptions. We then downloaded the protobuf formatted
tree (tree.all\_viridian.202409.pb.gz), exported the mutations to 
nucleotide format using \texttt{matUtils summary --translate}
(usher version 0.6.3 from bioconda), and translated
these mutations into tskit site and mutation tables. 
The resulting file 
 contains 5,345,019 nodes, 3,364,841
mutations and requires 29MB of storage space in tszip format (usher\_viridian\_v1.0.trees.tsz).
We validated that the data encoding by comparing the 
variant data for the UShER tskit tree sequence against our MAFFT alignments.
There was an exact match between the non-ambiguous nucleotide calls 
at 27,469 sites. 
% Usher tree has 27,508 vs sc2ts 29,893, or 2385 fewer sites 
At the remaining 39 sites for which the UShER tree 
had mutations there were a significant 
number of gap characters (deletion calls), suggesting that the mismatches
could be due to differences in the alignments used as input to UShER.
See notebook
\href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/validate\_usher\_tree.ipynb}{validate\_usher\_tree.ipynb}
for details.

We then computed intersection ARGs for
sc2ts (sc2ts\_viridian\_inter\_v1.2.trees.tsz) and
UShER (usher\_viridian\_inter\_v1.2.trees.tsz)
by simplifying \cite{Kelleher2018efficient,Wong2024}
both to the 2,475,418 samples and 27,507 sites that are shared
(the UShER tree has mutations at 27,508 sites and sc2ts ARG has 29,893).
As there was no documented means of deriving node dates from
the UShER tree, and we wish to visually compare
the phylogenetic backbones,
we also dated the intersection tree by fixing
the sample dates using the ``Date\_tree'' metadata field,
and using tsdate (version 0.2.3) to date internal nodes.

@@

\subsubsection*{Comparison of phylogenetic backbones}
We compare the non-recombinant parts of the ARG with the UShER topology
by simplifying the phylogenies to a smaller number of
Pango representative samples.
We focus on the 1473 Pango lineages that have perfect ``origination events'' 
(Document \ref{sec:supplement_pango_lineages}). We then omit
omit X lineages, leaving 1329 lineages.
To ensure that each sample is a tip in the simplified topologies,
a representative sample is then chosen as the oldest sample
descending from the origination node
whose own descendants (if any) are entirely of the focal Pango type.
We then further remove 4 lineages
which are singleton descendants of a recombination event in the ARG,
% NOTE: would have been simpler to say we just got rid of XBB and 
% descendants.
as well as 43 samples that are descendants of the recombinant XBB
but which do not start with an ``X''.
This results in a simplified ARG of 1282 samples,
each sample from a different Pango lineage.

The simplified ARG contains only two recombination nodes, which
are the two likely artefactual events discussed in 
Document \ref{sec:spurious-ancestrally-important-recombinants}, 
both of which have breakpoints near the 5' end of genome.
We therefore compare the Usher topology against the first tree in the ARG (covering
positions 1-26858; the majority of the genome). To reduce plot density, we consider
only those Pango lineages with greater than 10 samples.
Figure \ref{fig:tanglegram_base} shows the resulting tree compared with 
the equivalent dated UShER tree as a tanglegram.
We also show subtrees for Delta (Figure~\ref{fig:tanglegram_delta}),
BA.2 (Figure~\ref{fig:tanglegram_ba2}),
and BA.5 (Figure~\ref{fig:tanglegram_ba5}).

The order of tips in each plot was chosen to reduce tangling,
as calculated using the NeighborNet algorithm from 
Dendroscope \cite{Scornavacca2011,Huson2012},
with tanglegrams plotted using the SVG outputting facility built into tskit.
See \href{https://github.com/jeromekelleher/sc2ts-paper/blob/main/notebooks/supp\_pdf-cophylogeny-tanglegrams.ipynb}{supp\_pdf-cophylogeny-tanglegrams.ipynb}
for details.
